D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-01 ootavacuous.tex
Rework real-world constraints, mainly from Jade's questions
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@

The reason that atomic stores cannot be invented is that doing so can
introduce new (and almost certainly undesireable) behaviors.
To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
a CPPMEM\footnote{
	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
litmus test that demonstrates such a behavior.
Without line~4, only the values~2 and~3 can be stored to \co{r1}.
With that line, the additional value~42 can also be stored to \co{r1}.
The compiler is therefore forbidded from inventing that store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
For example, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

The speed of light is finite~\cite{OleRoemer1671SpeedOfLight}
and atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
These laws of physics mean that if one thread loads the value
stored by some other thread, that load must have occurred later
in global time than did the store.
In other words, the reads-from relation is temporal in nature.

In contrast, the C++11 modification-order relation is atemporal.
This means that the store that executed earlier in global time
might appear later in the stored-to object's modification
order~\cite{McKenney20xxParallelProgramming}.
The atemporal nature of C++11 modification order is due to the hardware
store-buffer optimizations used by modern multicore systems.

Heavily used CPU families now have accurate and executable formal
memory models, but even more important, there is much more clarity
on exactly what hardware will and will not do in the course of
speculative execution.

For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{a[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
	nor cares.}
Then line~7 can cause the final value of \co{r1} to be 42, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

Therefore, non-speculative loads are prohibited from loading values from
a speculative stores.

In addition, in order to avoid hardware-based OOTA, a value stored
speculatively will not be exposed to loads from threads other than the
one executing that store.

@@

\section{Wording}
\label{sec:Wording}

@@

% Done:
% David Goldblatt
% Boqun Feng?
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  Peter Sewell?  Mark Batty? Other TFWM authors?
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.

>>>>

The following sections discuss these constraints, starting with the
laws of physics, continuing with hardware constraints, and ending
with constraints imposed by the standard.

Evidence to date suggests that the universe is
causal~\cite{Plato360BC-causality},
the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
These fundamental laws of physics are empirical in nature, but
are backed by a great many observations extending back to a time
preceding electronic computers, let alone computer languages that
support concurrency.

The relationship between these three laws of physics and OOTA
cycles is described in the following sections.

\subsubsection{Causality}
\label{sec:Causality}

Plato's ``Timaeus'' notwithstanding, there has been much recent debate
as to whether the universe is in fact causal with cause necessarily
preceding effect.
However, no one has managed to construct a real time-travel machine
or any other apparatus that might result in effect preceding cause,
as discussed at length in the Wikipedia article on retrocausality and
its many references.\footnote{
	\url{https://en.wikipedia.org/wiki/Retrocausality}.}

This paper will therefore assume that effect cannot precede cause in
any real-world computing system.

\subsubsection{Finite Speed of Light}
\label{sec:Finite Speed of Light}

If the speed of light was infinite, then information could be transmitted
at infinite speed, arriving at any destination, no matter how remote,
in zero time.
This situation would permit a load to return the value stored by some
other thread at the exact instant that the store executed, which
would in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

However, all macroscopic measurements to date have shown the speed of
light to be finite, thus preventing causal loops based on infinite-speed
propagation of information.

This paper will therefore assume that the speed of light is finite, and
thus the maximum speed at which information can propagate is also finite.

\subsubsection{Non-Zero Sized Atoms}
\label{sec:Non-Zero Sized Atoms}

If atoms were of zero size, a zero-sized computer might be constructed.
In such a computer, even give finite speed of light, information could
propagate among computing elements in zero time.
This situation would also permit a load to return the value stored by
some other thread at the exact instant that the store executed, which
would again in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

However, there has to date been no successful attempt to construct a
zero-sized computing device capable of running non-trivial C++ programs.

This paper will therefore assume that any computing device capable of
running C++ will be of non-zero size.

\subsubsection{Laws of Physics: Consequences}
\label{sec:Laws of Physics: Consequences}

These three laws of physics mean that if one thread loads the value
stored by some other thread, that load must have occurred later
in global time than did the store.
In other words, the reads-from relation is temporal in nature.

In contrast, the C++11 modification-order relation is atemporal,
because computing systems can and observably do determine the
modification order of a set of concurrent stores long after the fact.
This means that the store that executed earlier in global time
might appear later in the stored-to object's modification
order~\cite{McKenney20xxParallelProgramming}.
The atemporal nature of C++11 modification order is due to the hardware
store-buffer optimizations used by modern multicore systems.

In happy contrast to the situation at the start of the C++11 memory-model
effort, heavily used CPU families now have accurate and executable formal
memory models, all of which prohibit the causal cycles required to form
OOTA cycles.
Even more important, there is clarity on exactly what hardware can and
can not do in the course of speculative execution.
Finally, computation is based on instructions, and instructions require
finite time to execute, even when executed speculatively.

For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{a[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
	nor cares, nor should it.}
Then line~7 can cause the final value of \co{r1} to be 42, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

Therefore, non-speculative loads are prohibited from loading values from
speculative stores.

As a special case of this prohibitino, a value stored speculatively must
not be exposed to non-speculative loads from other threads.
In the further special case of multiple hardware threads sharing a core,
one might imagine speculation involving multiple threads.
However, in this case, any squashing of speculation must squash the
full extent of that speculation, across all of the hardware threads
that were involved.

Finally, the prohibition of hardware OOTA applies within a single
multi-threaded core as well as between cores.
This means that for any set of speculative accesses to be committed,
all loads in that set must have been confirmed by corresponding
non-speculative stores.

@@

The reason that atomic stores cannot be invented is that doing so can
introduce new (and almost certainly undesirable) behaviors.
To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
a CPPMEM\footnote{
	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
litmus test that demonstrates such a behavior.
Without line~4, only the values~2 and~3 can be stored to \co{r1}.
With that line, the additional value~42 can also be stored to \co{r1}.
The compiler is therefore forbidded from inventing that store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
For example, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

@@

\section{Suggested Changes to Wording}
\label{sec:Suggested Changes to Wording}

@@

% Done as of November 30, 2023:
% David Goldblatt
% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.
